---
title: "fairgwr: introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fairgwr: introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Abstract
This vignette shows the basic usage of the `fairgwr` package.


# Introduction
Loading our package:

```{r setup}
library(fairgwr)
```

## Dataset
For this example we'll use the included `camden` dataset given its small size.

```{r}
#data(camden)
data(sevilla)
```

If needed, the preprocessing routine can generate datasets for any town in
the Spanish region of Andalusia. This can be checked in its own vignette:

```{r eval=FALSE}
vignette("preprocessing", package = "fairgwr")
```

Alternatively, an user-prepared dataset can be used given that the following
are provided:

* `x`: an $X_{(n\times p+1)}$ design matrix, where $\boldsymbol{1}$ is the first column. If
  $\beta_0$ is not needed, an $X_{(n\times p)}$ matrix can be provided
* `y`: an $Y_{(n\times 1)}$ matrix containing the observed values of the response
  variable
* `d`: a $D_{(n \times n)}$ distance matrix
* (optional) a `geometry` object containing the coordinates of every row in $X$,
  usually as an `sfc` object. Only needed if we plan on plotting the results

# Running

## Determine the sensitive variable
In the cases we don't have a explicitly defined variable to work with, we can
get the $u(x_j,y)$ measure of variation for every variable like:

```{r}
#databias(camden$x, camden$y)
databias(sevilla$x, sevilla$y)
```

For `camden`, `elder_ppl` (i.e. the $3$rd column) is the variable with the
highest percentage of variation and so will be our sensitive variable. Usually,
the value splitting the sensitive and non-sensitive classes is the median for 
the variable, but any other can be chosen if we so desire:

```{r}
#camden$split_col <- 3
#camden$split_val <- median(camden$x[, camden$split_col])
sevilla$split_col <- 4
```

A `classes` object will be necessary moving forward. This is just a list storing
which elements of $X$ belong to each class, which will come in handy later. This
can be done easily:

```{r}
sevilla$classes <- fair.classes(x = sevilla$x,
                                y = sevilla$y,
                                d = sevilla$d,
                                select_col = sevilla$split_col)
names(sevilla$classes)
```

## Choosing a fairness level
In our problem, we want to reduce the difference between the means of the
predicted values in the sensitive class minus the ones in the non-sensitive
class. This can be expressed as:
$$
    g\left(\boldsymbol{\beta}\right)=
     \frac{1}{card\left(S\right)}X_{S}\hat{\beta}_{S} -
     \frac{1}{card\left(S^C\right)}X_{S^C}\hat{\beta}_{S^C}
$$
where
$$
    S=\text{subset of elements in the sensitive class}
$$
briefly, this can be written:
$$\left\lvert g\left(\boldsymbol{\beta}\right)\right\rvert\leq\varepsilon$$
This will be one of the parameters driving the optimization problem. As such,
we can obtain the reference $\varepsilon$ value, that is, the current bias
present in the data:

```{r}
#fair.epsilon(camden$classes)
fair.epsilon(sevilla$classes)
```

Even though the value of $\varepsilon$ is automatically calculated by the `gwr`
routines, it is useful to store the unfairness reduction percentage we want for
futures use:

```{r}
# 10% reduction
fair.perc <- 0.1
# store epsilon value
#e10 <- fair.epsilon(camden$classes, fair.perc)
e10 <- fair.epsilon(sevilla$classes, fair.perc)
e10
```

## Calculate the optimal bandwidth
The first part of the optimization problem is the one in which the optimal
bandwidth (or $h$) is chosen. This is done by minimizing the CV-score function:
$$
\begin{split}
    CVSS(h)&=\sum_i \left( y_i - \check{y}_i(h)\right)^2 \\
    s.t.\quad&\left\lvert g(\boldsymbol{\beta}(h))\right\rvert\leq\varepsilon
\end{split}
$$
In our case, when running this problem, we also want the optimal $h$ to meet the
fairness requirement we defined earlier. How to do this:

```{r}
sevilla$h <- gwr.h_sel(x = sevilla$xnorm,
                       y = sevilla$y,
                       d = sevilla$d,
                       wfun = wmat.bisquare,
                       fair.perc = fair.perc,
                       classes = sevilla$classes,
                       constrained = TRUE,
                       verbose = FALSE)$h
```

In this step, a _weighting function_ is used. In our case, bisquare weights
(as `wmat.bisquare`) are used:
$$w_{ij}=\begin{cases}
    \left( 1-(d_{ij}/h)^2\right)^2 & \text{if } d_{ij} < h, \\
    0 & \text{otherwise.}
\end{cases}$$
Gaussian weights are also available under the `wmat.gauss` function. If needed,
any other weighting function can be easily adapted.

## Obtaining GWR results
The results of running a constrained geographically weighted regression can
be obtained as:

```{r}
results <- gwr(x = sevilla$xnorm,
               y = sevilla$y,
               d = sevilla$d,
               h = sevilla$h,
               wfun = wmat.bisquare,
               fair.perc = fair.perc,
               classes = sevilla$classes,
               constrained = TRUE)
```

Which will run the "global" GWR solver. Depending on the machine, this can take
a few minutes.

### Constrained solution
Provided that we are running `gwr` with `constrained = TRUE`, the constrained
solution will reflect our requested unfairness reduction. The `solution` and
`pred` attributes will give us the $\boldsymbol{\beta}$ matrix and the estimated
$\hat{y}$ values, respectively:

```{r}
head(results$solution)
head(results$pred)
```

It can be shown that this solution matches the fairness level we established
earlier using the $u$ measure of fairness:
```{r}
u10 <- fair.u(x = sevilla$xnorm,
              classes = sevilla$classes,
              betas = results$solution)
u10
e10
```

### Unconstrained solution
The unconstrained solution (i.e. regular GWR) is also returned when running
`gwr`. See the `unconstrained.solution` and `unconstrained.pred`, returning
$\boldsymbol{\beta}$ and $\hat{y}$ for the unconstrained problem.

```{r}
head(results$unconstrained.solution)
head(results$unconstrained.pred)
```

This behavior can also be achived when running `gwr` with `constrained=FALSE`.
In this case, none of `classes` and `fair.perc` are needed.

## Plotting
Once `gwr` is done running, we can generate maps to visualize our results if
we so desire. The first thing to do is generate a `basemap`, which is just
a map for the contour of our region of interest. This can be done as:

```{r, out.width="80%", fig.align = 'center', fig.retina = 3}
sevilla$basemap <- gwr.basemap(sevilla$geometry)
sevilla$basemap
```

The most interesting and useful map is probably the one for the predicted
values:

```{r, out.width="80%", fig.align = 'center', fig.retina = 3}
sevilla$predmap <- gwr.predmap(basemap = sevilla$basemap,
                               geometry = sevilla$geometry,
                               y = results$pred,
                               color.scales = list(highs = "Greens",
                                                  lows = "YlOrBr"),
                               color.rev = TRUE) +
                      ggplot2::theme(legend.position = "none")
sevilla$predmap
```

Which can be compared with the original $y$ values:

```{r, out.width="80%", fig.align = 'center', fig.retina = 3}
gwr.predmap(basemap = sevilla$basemap,
            geometry = sevilla$geometry,
            y = sevilla$y,
            color.scales = list(highs = "Greens",
                               lows = "YlOrBr"),
            color.rev = TRUE) +
    ggplot2::theme(legend.position = "none")
```

Maps for the values of the $\boldsymbol{\beta}$ matrix can also be generated,
for the case we want visualize their variation based on where they are located:

```{r, out.width="80%", fig.align = 'center', fig.retina = 3}
sevilla$betamaps <- gwr.betamaps(basemap = sevilla$basemap,
                                 geometry = sevilla$geometry,
                                 betas = results$solution)
# arrange plots in grid, removing legends and axes
gridExtra::grid.arrange(
    sevilla$betamaps[[1]] + ggplot2::theme_void() + ggplot2::theme(legend.position = "none"),
    sevilla$betamaps[[2]] + ggplot2::theme_void() + ggplot2::theme(legend.position = "none"),
    sevilla$betamaps[[3]] + ggplot2::theme_void() + ggplot2::theme(legend.position = "none"),
    sevilla$betamaps[[4]] + ggplot2::theme_void() + ggplot2::theme(legend.position = "none"),
    ncol = 2
)
```
